NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-21 Manuscript/manuscriptPS.tex
Update text

To evaluate counterion binding in different simulation against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{The upcoming results from Lipid17 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1522151331118\#c5876812436739342658} \\
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}
In line with the trend observed for bilayers in the absence of negatively
charged lipids \cite{??}, the POPC headgroup order parameter response to lithium ions is
significantly larger than to potassium, indicating stronger binding of lithium
ions to lipid bilayers. The unique nature of lithium is also evident in the
response of PS headgroup order parameters to bounds ions, which decrease
with the addition of lithium or divalent cations but increase with the
addition of sodium or potassium. The headgroup order parameter response
to the added NaCl in the Berger model is similar to the experiments of LiCl,
except for the $\alpha$ carbon of POPS for which the change is overestimated
with the order of magnitude. The results suggests that the binding affinity
of sodium is overestimated by the Berger model, in line with the previous results \cite{catte16},
and that the response of $\alpha$ carbon of POPS to the added sodium
is significantly overestimated. The response of POPC to the added potassium
in MacRog model is only slightly overestimated, in contrast to the response
of POPS where order of magnitude overestimations of reposonse are observed. 
In conclusion, it seems that the response of PS headgroup to bound cations
is not correctly captured in the tested models.
\todo{Finish the discussion when more models are available}

@@

\subsection{Headgroup structure in PS and PC mixtures}
Cellular membranes often compose of mixtures of zwitterionic and negatively
charged lipids. The response of PC headgroup to mixtures of differently charged
lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
As expected from the electrometer concept \cite{seelig87}, the headgroup 
order parameters increase with the addition of negatively charged PS and PG lipids,
decrease when mixed with positively charged surfactants and
are less affected by the addition of zwitterionic PE lipids or cholesterol.
In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
follow the eletrometer concept \cite{scherer87}.

Fig. \ref{HGorderparametersPCvsPS} shows the PC and PS headgroup order parameters
from POPC:POPS mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The experimentally observed increase of PC headgroup order parameters with the increasing
amount of negatively charged PS lipid is reproduced in
MacRog simulations with potassium counterions, but not in CHARMM36 simulations
with potassium or sodium conterions. The results can be
understood from the counterion density distributions shown in Fig. \ref{CIdensPSOCmixt}.
Both counterions are more strongly bound to bilayers in
CHARMM36 simulations than potassium in MacRog simulation.
The cation binding probably neutralizes the effect of negatively
charged lipids in CHARMM36 simulations, while the
weaker binding of potassium in MacRog simulations enables the
increase of order parameters with the added amount of PS.
\todo{This discussion will be finalized when we have the
results for monovalent ions also from other simulations.}

%$\beta$-carbon of POPC decreases and almost negligible increase is 
%seen in $\alpha$ order parameter with the addition of PS lipid 
%This may indicate incorrect interactions between PC and PS headgroups
%in simulations. On the other hand, the previously reported overestimated
%ion binding affinity may also play a role, as the amount of counterions
%is increasing with PS.
%\todo{Data from other force fields to be collected before discussion can be finalized.}

@@

\section{PC lipid headgroup response to different mixtures in experiments}
As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

>>>>

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}

@@

\subsection{Headgroup structure in PS and PC mixtures}
Dilution of PS lipid bilayers with PC lipids reduces propensity for the formation of
strong complexes with multivalent ions and is proposed to make PS headgroups 
less rigid~\cite{browning80,buldt81,roux90,roux91}. Therefore, the intermolecular
interactions at the headgroup region seems to determine important physical properties 
of mixed lipid bilayers. The headgroup order parameters of POPC increase in experiments 
with increasing amount of POPS in the mixed lipid bilayer (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
This behaviour is generally observed when negatively charged lipids or surfactants 
are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
tilting of lipid headgroup more parallel to the membrane plane according to
the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

The increase of POPC headgroup order parameters with the increasing
amount of negatively charged POPS lipid is reproduced in
MacRog simulations with potassium counterions, but not in 
Berger simulations with sodium or in CHARMM36 simulations
with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
The observations can be explained using the electrometer concept.
The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
which surpasses the effect of negatively charged lipids as the amount of
counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Finally, the weak binding of potassium in the MacRog simulations
enables the increase of order parameters with the added amount of PS 
(Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).

@@
